% This file originated as
% phscore:/home/jgrab/AnalysisSoftware/Intertwolog/Intertwolog5.1/README.
% Accessed 07-Nov-2014 by David McWilliams.  It was the latest version readme
% found under the Intertwolog directory tree.  The file time stamp was Jul 18,
% 2008. Simple reformatting has been applied.

\section{Intertwolog}
\label{sec:i2l}
\subsection{Description}
Intertwolog reads genotype data from a genotype file and affection status data
from a phenotype file and calculates the two-way interaction beta, interaction
beta p-value and standard error for each combination of two markers.  The
two-way interaction calculation is set up as an additive logisitic regression
model with an interaction term corrected for colinearity.  Only the beta,
p-value and standard error for the interaction term are written to the output
file.

\subsection{Usage}
\label{sub:i2l_usage}
See ``SNPlash'', p.\pageref{sec:snplash}, for options common to all engines.

\begin{verbatim}
  snplash -engine intertwolog -geno <filename> -phen <filename> \
          -out <filename> -map <filename>
\end{verbatim}

\subsection{Output}
A header section is printed with information about the files read and options
used.  The variable labels are multi-line.  The values printed are:

\begin{description}
\item[Marker 1 Position:] The genotype file position of the first marker used
  for the two way interaction calculations.

\item[Marker 1 Name:] The SNP name of the first marker used for the two way
  interaction calculations.  Appears in the output file if the map file option
  is used.

\item[Marker 2 Position:] The genotype file position of the second marker used
  for the two way interaction calculations.

\item[Marker 2 Name:] The SNP name of the second marker used for the two way
  interaction calculations.  Appears in the output file if the map file option
  is used.

\item[Interaction Beta] The logistic regression interaction beta.

\item[Interaction P Value] The p-vlaue for the logistic regression interaction beta.

\item[Interaction Standard Error] The Standard Error for the logistic regression
  interaction beta.
\end{description}

\noindent{}Example output:
\begin{verbatim}
Marker 1 Marker 1 Marker 2  Marker 2 Interact.  Interact.  Standard  
Position Name     Position  Name     Beta       P Value    Error     
-------- -------- --------  -------- ---------- ---------- ----------
       1 Marker01        2  Marker02   -0.42848    0.08704  0.25039
       1 Marker01        3  Marker03   -0.18137    0.38484  0.20870
       1 Marker01        4  Marker04   -0.12193    0.53099  0.19462
       1 Marker01        5  Marker05   -0.14906    0.63916  0.31792
\end{verbatim}

%% End intertwolog.tex \\
